---
title: "Assignment 8: Time Series Analysis"
author: "Vanshika Mittal"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A08_TimeSeries.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

## Set up 
1. Set up your session:

* Check your working directory
* Load the tidyverse, lubridate, zoo, and trend packages
* Set your ggplot theme

```{r}
getwd() #working directory is set 
#install.packages("trend")
#loading necessary packages 
library(tidyverse); library(lubridate); library(zoo); library(trend)

#setting my theme 
my_theme <- theme_minimal()+
  theme(plot.background = element_rect ('white'),
        legend.position = "top", 
        plot.title = element_text(color = 'black'),
        legend.box.background = element_rect(color = "black"))
theme_set(my_theme)
```


2. Import the ten datasets from the Ozone_TimeSeries folder in the Raw data folder. These contain ozone concentrations at Garinger High School in North Carolina from 2010-2019 (the EPA air database only allows downloads for one year at a time). Import these either individually or in bulk and then combine them into a single dataframe named `GaringerOzone` of 3589 observation and 20 variables. 

```{r, message = FALSE}
#1 importing all datasets 
O3_10 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2010_raw.csv",
                       header = TRUE, stringsAsFactors = TRUE)
O3_11 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2011_raw.csv",
                       header = TRUE, stringsAsFactors = TRUE)
O3_12 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2012_raw.csv",
                       header = TRUE, stringsAsFactors = TRUE)
O3_13<- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2013_raw.csv",
                       header = TRUE, stringsAsFactors = TRUE)
O3_14 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2014_raw.csv",
                       header = TRUE, stringsAsFactors = TRUE)
O3_15<- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2015_raw.csv",
                       header = TRUE, stringsAsFactors = TRUE)
O3_16 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2016_raw.csv",
                       header = TRUE, stringsAsFactors = TRUE)
O3_17 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2017_raw.csv",
                       header = TRUE, stringsAsFactors = TRUE)
O3_18 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2018_raw.csv",
                       header = TRUE, stringsAsFactors = TRUE)
O3_19 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2019_raw.csv",
                       header = TRUE, stringsAsFactors = TRUE)
#combining the datasets via their rows 
GaringerOzone <- rbind(O3_10, O3_11, O3_12, O3_13, O3_14, O3_15, O3_16, O3_17,
                       O3_18, O3_19)
#checking the dimensions
dim(GaringerOzone)
```

## Wrangle

3. Set your date column as a date class.

4. Wrangle your dataset so that it only contains the columns Date, Daily.Max.8.hour.Ozone.Concentration, and DAILY_AQI_VALUE. 

5. Notice there are a few days in each year that are missing ozone concentrations. We want to generate a daily dataset, so we will need to fill in any missing days with NA. Create a new data frame that contains a sequence of dates from 2010-01-01 to 2019-12-31 (hint: `as.data.frame(seq())`). Call this new data frame Days. Rename the column name in Days to "Date". 

6. Use a `left_join` to combine the data frames. Specify the correct order of data frames within this function so that the final dimensions are 3652 rows and 3 columns. Call your combined data frame GaringerOzone. 

```{r}
# 3 changing the date column 
GaringerOzone$Date <- mdy(GaringerOzone$Date)

# 4 keeping a few columns 
GaringerOzone <- select(GaringerOzone, Date, Daily.Max.8.hour.Ozone.Concentration,
                        DAILY_AQI_VALUE)

# 5
start_date <- as.Date("2010-01-01") #assigning a start date 
end_date <- as.Date("2019-12-31") #assigning an end date 
#creating a new df with a sequence of dates from our start to end date 
Days <- as.data.frame(seq(start_date, end_date, by = "days")) 
colnames(Days) <- "Date" #renaming the column 

# 6 joining the date sq. df to find NA values 
GaringerOzone <- left_join(Days, GaringerOzone)
dim(GaringerOzone) #checking dimensions of our new df 
```

## Visualize

7. Create a line plot depicting ozone concentrations over time. In this case, we will plot actual concentrations in ppm, not AQI values. Format your axes accordingly. Add a smoothed line showing any linear trend of your data. Does your plot suggest a trend in ozone concentration over time?

```{r}
#7
Ozone_ppm <- ggplot(GaringerOzone, aes(x = Date, 
                                       y = Daily.Max.8.hour.Ozone.Concentration)) + 
  geom_point(color = "lightblue") + 
  geom_smooth(method = lm, color = "blue") +
  labs(x = "Years", y = "Ozone Concentration (ppm)")

Ozone_ppm 
```

>Answer: Over the years, there is a slight decrease in the level of ozone concentrations. 

## Time Series Analysis

Study question: Have ozone concentrations changed over the 2010s at this station? 

8. Use a linear interpolation to fill in missing daily data for ozone concentration. Why didn't we use a piecewise constant or spline interpolation?

```{r}
#8 checking the number of NA's 
summary(GaringerOzone$Daily.Max.8.hour.Ozone.Concentration) 

#replacing NA's through linear interpolation 
GaringerOzone$Daily.Max.8.hour.Ozone.Concentration <-
  zoo::na.approx(GaringerOzone$Daily.Max.8.hour.Ozone.Concentration)

#checking whether NA's have been removed and to cross check the summary statistics for this variable
summary(GaringerOzone$Daily.Max.8.hour.Ozone.Concentration) 
#yes the NA's are gone and the summary statistics have not changed. 

```

> Answer: The spline interpolation uses a quadritic function which we do not need since we have a linear function. We could use either piecewise constant or the linear interpolation in this. 

9. Create a new data frame called `GaringerOzone.monthly` that contains aggregated data: mean ozone concentrations for each month. In your pipe, you will need to first add columns for year and month to form the groupings. In a separate line of code, create a new Date column with each month-year combination being set as the first day of the month (this is for graphing purposes only)

```{r}
#9 finding mean ozone concentrations each month 
GaringerOzone.monthly <- 
  GaringerOzone %>% 
  mutate (Year = year(Date), Month = month(Date)) %>% #separating month and year 
  group_by(Year, Month) %>% #grouping by month and year 
  summarise(mean_ozone_conc = mean(Daily.Max.8.hour.Ozone.Concentration)) 

#creating a new column with month, year and first day of each month 
GaringerOzone.monthly <- 
  GaringerOzone.monthly %>% 
  mutate(Date = paste(Year,"-" ,Month,"-",01))
#assigning it a date object 
GaringerOzone.monthly$Date <- ymd(GaringerOzone.monthly$Date)
```


10. Generate two time series objects. Name the first `GaringerOzone.daily.ts` and base it on the dataframe of daily observations. Name the second `GaringerOzone.monthly.ts` and base it on the monthly average ozone values. Be sure that each specifies the correct start and end dates and the frequency of the time series.

```{r}
#10 assigning the start dates from the dataset 
f_year = 2010
f_month = 1
f_day = 1
# generating time series object from the daily dataset 
GaringerOzone.daily.ts <- ts(GaringerOzone$Daily.Max.8.hour.Ozone.Concentration, 
                             start= c(f_year, f_month, f_day), frequency = 365)

#generating time series object from the monthly dataset 
GaringerOzone.monthly.ts <- ts(GaringerOzone.monthly$mean_ozone_conc, 
                             start= c(f_year, f_month), frequency = 12)
```

> Please Note: We do not specify the end date in this code as we have already corrected for the missing values. This means that R will automatically stop at the end of the dataset. 

11. Decompose the daily and the monthly time series objects and plot the components using the `plot()` function.

```{r}
#11 decomposing daily data time series object 
ozone.daily.decomp <- stl(GaringerOzone.daily.ts, s.window = "periodic")
plot(ozone.daily.decomp)

#decomposing monthly data time series object 
ozone.monthly.decomp <- stl(GaringerOzone.monthly.ts, s.window = "periodic")
plot(ozone.monthly.decomp)
```

12. Run a monotonic trend analysis for the monthly Ozone series. In this case the seasonal Mann-Kendall is most appropriate; why is this?

```{r}
#12
#H0: the dataset is stationary 
#Ha: the dataset has a trend 
ozone.monthly.trend <- Kendall:: SeasonalMannKendall(GaringerOzone.monthly.ts)
summary(ozone.monthly.trend)
#reported p-value is less than 0.5, we can reject the null hypothesis 
```

> Answer: We run the the seasonal Mann-Kendall test on the monthly Ozone series since we have just confirmed (in step 11) that the monthly ozone series follows a seasonal trend. 

13. Create a plot depicting mean monthly ozone concentrations over time, with both a geom_point and a geom_line layer. Edit your axis labels accordingly.

```{r} 
# 13
ozone.monthly.trend.plot <- 
  ggplot(GaringerOzone.monthly, aes(x=Date, y = mean_ozone_conc)) + 
  geom_point() + 
  geom_line()+
  labs(x= "Time", y = "Monthly Mean Ozone Concentrations") + 
  geom_smooth(method = lm)

ozone.monthly.trend.plot
```

14. To accompany your graph, summarize your results in context of the research question. Include output from the statistical test in parentheses at the end of your sentence. Feel free to use multiple sentences in your interpretation.

> Answer: H0: The ozone concentration has not changed over 2010's at this station. 
Ha: The ozone concentration has changed over 2010's at this station
Along with the trend that can be observed in the graph and from the statistical output {p-value = 0.0467} which is less than 0.5, we can reject the null hypothesis and say that the ozone concentration has changed over 2010s. 

15. Subtract the seasonal component from the `GaringerOzone.monthly.ts`. Hint: Look at how we extracted the series components for the EnoDischarge on the lesson Rmd file.

16. Run the Mann Kendall test on the non-seasonal Ozone monthly series. Compare the results with the ones obtained with the Seasonal Mann Kendall on the complete series.

```{r}
#15 decomposing the components and then combining trend and remainder back 
GaringerOzone.monthly.Components <- 
  as.data.frame(ozone.monthly.decomp$time.series[,2:3])

nonseasonal.GaringerOzone.monthly <- (GaringerOzone.monthly.Components$trend 
                                +GaringerOzone.monthly.Components$remainder)

#assign a new ts object 
nonseasonal.GaringerOzone.monthly.ts <- ts(nonseasonal.GaringerOzone.monthly, 
                                    start= c(f_year, f_month), frequency = 12)

#16 conducting the test 
nonseasonal.GaringerOzone.monthly.ts.trend <- 
  Kendall::MannKendall(nonseasonal.GaringerOzone.monthly.ts)
#summarising the results 
summary(nonseasonal.GaringerOzone.monthly.ts.trend)
```

> Answer: Our p-value is less than 0.5 which means we can reject the null hypothesis and concludde that the our data is not stationary and has a trend over the given period of time. These results are comparable to the Seasonal MannKendall test which I ran earlier as they both echo the same result. 
