---
title: "Assignment 2: Coding Basics"
author: "Vanshika Mittal"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.
```{r}
seq (from = 1, to = 30, by = 3) #generating a sequence 
```

2.  Compute the mean and median of this sequence.
```{r}
seq1 <- seq(from = 1, to = 30, by = 3) #assigning a sequence name 
mean(seq1) #calculating its mean and media 
median(seq1)
```

3.  Ask R to determine whether the mean is greater than the median.
```{r}
#assigning the mean and median of a sequence to variables 
mean.seq1 <- mean(seq1) 
median.seq1 <- median(seq1)
mean.seq1 > median.seq1 #determining whether mean is greater than median
```


4.  Insert comments in your code to describe what you are doing.
see codes above

```{r basics 1}
#1. 

#2. 

#3. 
```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.
```{r}
student_name <- c("Ally", "Ben", "Charlie", "Dina") #creating a vector with student names 
student_score <- c(88, 45, 67, 93) #creating a vector with random scores of choice 
passing_score <- c(TRUE, FALSE, TRUE, TRUE) #creating a vector based on the passing grade of 50
```

6.  Label each vector with a comment on what type of vector it is.
```{r}
#finding what class do the vectors belong to
class(student_name)
class(student_score)
class(passing_score)
```
student_name: character vector 
student_score: numeric vector 
passing_score: logical vector 

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.
```{r}
report_card <- cbind(student_name, student_score, passing_score) #combining the above vectors 
is.data.frame(report_card) #checking whether this is a data frame
report_card_df <- as.data.frame(report_card) #converting the vector into a data frame
is.data.frame(report_card_df) #confirming whether its been converted 
```

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
colnames(report_card_df) <- c("Name", "Score", "Passing Grade") #renaming column names 
report_card_df #viewing our df to confirm
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer:All values in a matrix belong to the same class however this dataframe contains three different classes of vectors. The "Name" column is a character vector, "Score" column is numeric vector and "Passing Grade" column is a logical vector

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement.

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
#generating a loop of ifelse conditions for the entire "Score" column in our df
for (x in report_card_df$Score){
  Passing_Status <- ifelse (x > 50, print(TRUE), print(FALSE))
}
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:Both option 'if' and 'else' and 'ifelse' should since they imply the same working inputs. 
I have chosen to use the ifelse statement as it is easier to consolidate a larger argument for a vector in it.  
